Repeated loss of variation in insect ovary morphology highlights the role of development in life-history evolution

### 2021-04-02 manuscript.Rmd
proof read revisions April 2021

The number of offspring an organism can produce is a key component of its evolutionary fitness and life-history. Here we perform a test of the hypothesized trade-off between the number and size of offspring using thousands of descriptions of the number of egg-producing compartments in the insect ovary (ovarioles), a common proxy for potential offspring number in insects. We find evidence of a negative relationship between egg size and ovariole number when accounting for adult body size. However in contrast to prior claims, we note that this relationship is not generalizable across all insect clades, and we highlight several factors that may have contributed to this size-number trade-off being stated as a general rule in previous studies. We reconstruct the evolutionary history of the nurse cell arrangement within the ovariole, and show that the diversification of ovariole number and egg size have both been largely independent of nurse cell presence or position within the ovariole. Instead we show that ovariole number evolution has been shaped by a series of transitions between variable and invariant states, with multiple independent lineages evolving to have almost no variation in ovariole number. We highlight the implications of these invariant lineages on our understanding of the specification of ovariole number during development, as well as the importance of considering developmental processes in theories of life-history evolution.

@@

The insect female reproductive system includes a pair of ovaries, each of which contains a number of ovarioles[@buning1994insect] (Fig 1a). Each ovariole consists of an anterior germarium containing the stem cell niche or resting oogonia, developing oocytes arranged in an ontogenic series from anterior to posterior, and a posterior connection to a common oviduct. The number of ovarioles varies across species[@hodin2009she], and can vary across individuals in a population[@honvek1993intraspecific], as well as between the left and right ovary within a single individual[@iwata1955comparative]. Therefore total ovariole number may be an even or odd integer for an individual female insect. In addition to variation in the number of ovarioles, the tissue morphology within ovarioles varies across insects, and has been classified into several modes of oogenesis based on the presence and position of special nutritive cells called nurse cells[@buning1994insect].

Here we compiled `r nrow(raw_ON)` records of ovariole number from across `r ON %>% distinct(order) %>% nrow()` orders, `r ON %>% distinct(family) %>% nrow()` families, and `r ON %>% filter(species != "") %>% distinct(genus,species) %>% nrow()` species of insects. We combined these data with published datasets of egg size[@church2019dataset], fecundity[@Gilbert2007;@gilbert2010parental], and body size[@rainford2016phylogenetic], to test hypotheses about \textcolor{red}{the evolutionary trade-off between offspring size and number}. In these analyses we used an existing phylogeny of insects[@church2019insect] to analyze evolutionary patterns in ovariole number, and found that hypotheses about \textcolor{red}{life-history} evolution do not hold generally true across insects. \textcolor{red}{We then combined these data with published observations of the mode of oogenesis}[@buning1994insect]\textcolor{red}{ and reconstructed the evolutionary history of the presence and position of nurse cells that contribute to the oocyte during oogenesis. We tested whether patterns in the distribution of ovariole number, egg size, or egg shape were driven by the evolution of nurse cells, and found no significant results.} Instead we observe that the phylogenetic distribution of ovariole number suggests a model where the developmental mechanisms that govern ovariole number have shifted between variable and invariant states several times over the course of insect evolution. Based on this finding, we propose that the developmental mechanisms used to establish ovariole number in well-studied insects such as _Drosophila melanogaster_ are unlikely to regulate ovariole number in all insects.

@@

We searched the published literature for references to insect ovariole number using a predetermined set of 131 search terms, entered into Google Scholar (scholar.google.com) between June and October of 2019. Each search term was \textcolor{red}{comprised of} an insect taxonomic group and the words "ovariole number". \textcolor{red}{The taxonomic groups used to search included} all insect orders, many large insect families, and taxonomic groups that are well-represented in the insect egg dataset[@church2019dataset]. \textcolor{red}{For each Google Scholar search}, we evaluated all publications in the first page of results (ten publications). For 61 search terms that had a large number of informative hits, significant representation in the egg dataset, or that corresponded to very speciose groups, we evaluated an additional 20 publications. The list of search terms is available in the supplementary file 'ovariole_number_search_terms.tsv'.

Using this approach, we gathered `r nrow(raw_ON)` records for ovariole number from `r raw_ON %>% distinct(publication) %>% nrow()` publications, listed in the supplementary file 'ovariole_number_bibliography.pdf'. We matched these records to additional taxonomic information using the software TaxReformer[@TaxReformer]. For all subsequent analyses, we excluded observations made in non-reproductive individuals from eusocial species (workers), as well as two observations that represented significant outliers and could not be validated using additional sources[@su2015testicular;@hernandez2017biology]. See supplementary methods section \ref{sec:gathering} for details.

\textcolor{red}{For records of ovariole number that reported intraspecific variation in ovariole number, we calculated the percent difference as follows: if ovariole number was reported as a range, percent difference was calculated as the $100 * ((max - min) / median)$; if ovariole number was reported as an average with deviations, percent difference was calculated as $100 * ((2 * deviation) / mean)$. When independent observations of ovariole number for a given species were available across multiple published records, we calculated the percent difference as the $100 * ((max - min) / median)$.}

@@

The analyses in this manuscript were performed using the insect phylogeny published in Church et al., 2019[@church2019insect], unless otherwise specified. \textcolor{red}{For regressions involving body size data that were reported as insect family-level averages,} we used the insect phylogeny published in Rainford et. al, 2014[@rainford2014phylogenetic]. Analyses of Drosophilidae ovariole number, egg size, and body size were performed using a phylogeny newly assembled for this study. See supplementary methods section \ref{sec:trees} for details.

@@

Analyses of evolutionary rate were performed using BAMM (version 2.5.0)[@rabosky2014automatic]. For this analysis, we calculated the average ovariole number (log~10~ transformed) for each genus present in the phylogeny (507 taxa). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select appropriate priors, and ran BAMM for the maximum number of generations ($2*10^{-9}$), sampling every 10^6^ generations. Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically. Running BAMM for the maximum possible number of generations and selecting the optimum burn-in (Fig. \ref{fig:BAMM-analysis-burnin}) resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See supplementary methods section \ref{sec:BAMM} for details.

@@

All phylogenetic regressions were performed using the maximum clade credibility (MCC) tree \textcolor{red}{(the tree with highest credibility score from the posterior distribution of the Bayesian analysis)}. We considered a relationship significant when the p-value was below the threshold 0.01. To assess the robustness of results to uncertainty in phylogenetic relationships, we also \textcolor{red}{repeated these analyses over the posterior distribution of phylogenetic trees} and report the number of regressions that gave a significant result (see Table \ref{tab:PGLS-posterior-egg-ON}).

For two comparisons, we validated that our tests had sufficient statistical power using the selected threshold by comparing the distribution of p-values from regressions of observed data to regressions of data simulated under alternative hypotheses. We compared the results of analyses of our observed to those based on simulated data to evaluate the likelihood of false positives (comparing to data \textcolor{red}{simulated} under no correlation) and false negatives (comparing to data simulated with strong correlation).

@@

Ovariole number varies by at least four orders of magnitude across \textcolor{red}{insect species} (Fig. 1b). We identify seven insect families with species that have been reported to have more than 1,000 total ovarioles, including several eusocial insects (e.g. queens of the termite species _Hypotermes obscuriceps_, Blattodea: Termitidae[@bugnion1912anatomie], and several ant species, Hymenoptera: Formicidae)[@robertson1995sperm;@schneirla1957comparison] and non-eusocial insects (e.g. the blister beetle _Meloe proscarabaeus_, Coleoptera: Meloidae)[@buning1979trophic]. We also find two independent lineages that have evolved to have only one functional ovariole: dung beetles in the tribe Scarabaeinae (Coleoptera: Scarabaeidae)[@richter1974ovariole], and grass flies in the genus _Pachylophus_ (Diptera: Chloropidae)[@meier1999ovoviviparity;@pollock1996viviparous]. In these insects one of the two ovaries presumably established during embryogenesis is reported to atrophy during development[@pollock1996viviparous;@pluot1979evolution], resulting in an asymmetric adult reproductive system. \textcolor{red}{We also evaluated intraspecific variation in ovariole number, and found that, for species for which it has been reported, the average percent difference number within species is between 10\% and 100\% of the median value (Fig. \ref{fig:intraspecific-variation}).}

@@

Ovariole number has been hypothesized to be negatively correlated with egg size[@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. This hypothesis is based on the predictions that (1) female reproduction is resource-limited, therefore egg size should trade off with egg number, and (2) ovariole number can serve as a proxy for egg number[@montague1981reproductive;@smith1974optimal]. We did not observe a significant negative relationship when comparing egg size and ovariole number across insect species (Fig. 2a, Table \ref{tab:PGLS-posterior-egg-ON}, p-value `r round(pgls_results_egg_ON_name$tTable[[8]],3)`, n=`r pgls_results_egg_ON_name$dims$N`). We also compared egg size and ovariole number, \textcolor{red}{combining data from species within the same genus to increase sample size,} and again did not observe a significant relationship (Fig. \ref{fig:PGLS-egg-ON-genus}, p-value `r round(pgls_results_egg_ON_genus$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus$dims$N`). \textcolor{red}{To verify this finding was not driven by the high ovariole numbers seen in the queens of some eusocial insects, we repeated this comparison excluding insects from families with eusocial representatives, with the same result (Fig. \ref{fig:PGLS-egg-ON-genus-noneusocial}, p-value `r round(pgls_results_egg_ON_genus_noneusocial$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus_noneusocial$dims$N`).}

Given that this predicted relationship is often conditioned on body size, \textcolor{red}{which is predicted to limit total potential reproductive investment}[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with data on insect adult body mass[@Gilbert2007;@gilbert2010parental;@gilbert2011insect] and length[@rainford2016phylogenetic]. When accounting for adult body mass, we observed a significant negative relationship between egg size and ovariole number (Fig. 2b, \ref{fig:PGLS-Gilbert-name}, p-value `r round(pgls_results_gilbert_genus$tTable[[8]],3)`, slope `r round(pgls_results_gilbert_genus$coefficients[[2]],3)`, n=`r pgls_results_gilbert_genus$dims$N`). To evaluate the robustness of this result, we repeated the analysis 1000 times, taking into account uncertainty in both the phylogeny and trait measurements. Out of 1000 regressions, `r sum(sapply(allometry_gilbert_genus,function(x){x$tTable[[8]]}) < 0.01)` indicated a significant negative relationship (Table \ref{tab:PGLS-posterior-egg-ON}). We performed the same comparison accounting for adult body length, and likewise observed a significant negative relationship (Fig. \ref{fig:PGLS-family-length}, p-value `r gsub(0,"<0.001",round(pgls_results_body_egg_ON_family$tTable[[8]],3))`, slope `r round(pgls_results_body_egg_ON_family$coefficients[[2]],3)`, n=`r pgls_results_body_egg_ON_family$dims$N`), supported by `r sum(sapply(allometry_body_egg_ON_family,function(x){x$tTable[[8]]}) < 0.01)` of 1000 repeated analyses (Table \ref{tab:PGLS-posterior-egg-ON}).

We further explored these results using two methods: First, to evaluate our findings against alternative evolutionary hypotheses, we compared these results to regressions based on simulated data. Our results showed that when considering body size, the slope of the regression of egg size and ovariole number is \textcolor{red}{more negative} than we would expect to observe by chance, as assessed by comparing to data simulated with no evolutionary correlation (Fig. \ref{fig:simulated}). However, for both adult body length and dry mass, the slope of the regressions on observed data are not within the range that would be expected under a strong negative correlation (slope of -1 in log-log space, Fig. \ref{fig:simulated}). This suggests the presence of a weak evolutionary relationship between ovariole number and egg size, when accounting for body size.

@@

In addition to the number of ovarioles, insect ovary morphology has been classified into several modes of oogenesis based on the presence and position of nutritive cells called nurse cells[@buning1994insect], (Fig. 4a). Egg formation in the well-studied species _D. melanogaster_ is an example of a meroistic oogenesis mode, meaning that its ovarioles contain nurse cells of germ line origin that are connected to developing oocytes via cytoplasmic bridges[@KingRobertC1970OdiD]. In insects with a polytrophic arrangement, these nurse cells are clonally related and immediately adjacent to each oocyte. An alternative arrangement is seen in telotrophic meroistic ovaries, where oocytes in each ovariole are connected to a common pool of nurse cells located in the germarium[@buning1994insect]. Meroistic ovaries are thought to have evolved from an ancestral panoistic mode, meaning they lack nurse cells[@buning1994insect]. Using a previously published set of descriptions of these oogenesis modes across insects[@buning1994insect], we reconstructed the evolutionary transitions between these states. Consistent with previous analyses[@buning1994insect], we found that the ancestral insect likely had panoistic ovaries (lacking nurse cells), with several independent shifts to both telotrophic and polytrophic meroistic modes, and at least two reversals from meroistic back to panoistic (Figs. 4b, \ref{fig:reconstruct-ovary}).

@@

Using the dataset compiled here and a previously published phylogeny of insects (Fig. 5a)[@church2019insect], we modeled the rate of evolutionary change in ovariole number (Figs. \ref{fig:arbutus}, \ref{fig:BAMM-analysis-convergence}, \ref{fig:BAMM-analysis-burnin}, \ref{fig:BAMM-analysis-results}). We observed substantial rate heterogeneity in the evolution of ovariole number (Fig. \ref{fig:BAMM-analysis-results}), meaning that for some lineages ovariole number has evolved rapidly where in others, ovariole number has evolved very slowly or not at all. The most striking example of this are the multiple lineages which have independently evolved invariant or near-invariant ovariole number across taxa (e.g. nearly all Lepidoptera have exactly eight ovarioles, Fig. 5b, \textcolor{red}{Lepidoptera are part of Amphiesmenoptera}, in cyan), from an ancestral variable state. These invariant lineages were identified by finding regions of the phylogeny that experience extremely low rates of ovariole number diversification (Figs. \ref{fig:BAMM-analysis-results}, \ref{fig:BAMM-anlaysis-mean-rates}). Using this approach, we found that invariant ovariole numbers have evolved at least nine times independently across insects, with several subsequent reversals from invariant to variable states (Fig. 5a).

@@

The distribution of ovariole numbers across insects is enriched for even numbers of total ovarioles (Fig. 5c). While many insects show asymmetries in the number of ovarioles between the left and right ovaries, all of the invariant lineages are symmetric (at 4, 6, 8, 10, 12, and 14 total ovarioles). \textcolor{red}{Additionally, for the insects identified as part of invariant lineages, none have any reported intraspecific variation in ovariole number.} Therefore, invariant lineages have near-zero variation when comparing between species, between individuals within a species, and between the left and right ovary within an individual.

Using these results, we propose a multi-rate model, where the rate of ovariole number evolution differs based on the evolution of a discrete trait representing invariant or variable status. We propose that the evolution of this discrete trait is governed by a model where the likelihood of transitions from a variable to an invariant state is negatively correlated with the current number of ovarioles. Here we demonstrate that a multi-rate Brownian motion model far outperforms a single rate model in fitting the data ($\Delta$ AICc `r round(BM1$AICc - BMS$AICc,2)`). \textcolor{red}{In addition, using a parametric bootstrap to evaluate model fit, we find evidence that processes beyond Brownian Motion processes are likely at play (Fig. \ref{fig:arbutus})}[@pennell2015model]\textcolor{red}{. We suggest that as researchers continue to develop non-Gaussian models for continuous trait evolution}[@blomberg2020beyond]\textcolor{red}{, those models will be useful for describing the evolution of ovariole number.}

@@

A frequently invoked life-history prediction is that, given a finite set of metabolic resources, organisms can either produce few offspring, each with high fitness, or many low-fitness offspring[@smith1974optimal;@lack1947significance;@stearns1992evolution]. In insects, egg size and ovariole number are often used as proxies for offspring fitness[@koch2014mother] and number[@david1970nombre;@bouletreau1982reproductive], respectively, and therefore it has been predicted that insects with more ovarioles lay smaller eggs than insects with fewer ovarioles[@hodin2009she;@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. Our results, using a dataset that spans `r nrow(raw_ON)` observations and that takes into account phylogenetic relationships, indicate that a generalized trade-off between insect egg size and ovariole number does not exist (Fig. 2).

@@

We suggest that considering the evolution of developmental processes that govern ovariole number specification may be more useful in explaining patterns of diversity than predictions based on metabolic trade-offs. \textcolor{red}{As evidence of this, we point to the fact that invariant lineages appear to have near-zero variation not only across species, but also within species, and between the left and right ovary within individuals. This suggests that the mechanism that determines ovariole number has become canalized in these groups. In contrast,} our previous understanding of how ovariole number is regulated comes from research on _Drosophila melanogaster_, where the number of ovarioles can vary between the left and right ovaries within an individual, as well as across individuals within a population[@telonis2005genetic;@bergland2008quantitative]. In this species, adult ovariole number is determined by cell proliferation and rearrangement during larval development[@godt1995mechanisms;@king1968development]. Variation in adult number is derived primarily from variation in the number of “terminal filament precursor cells”[@green2012convergent;@sarikaya2019reproductive], as well as from variation in the number of those precursor cells that group together to form the structure that initiates ovariole formation, known as a “terminal filament”[@sarikaya2012roles]. Across species of _Drosophila_, variation in average adult ovariole number results primarily from variation in the average number of terminal filament precursor cells[@sarikaya2019reproductive].

>>>>

The number of offspring an organism can produce is a key component of its evolutionary fitness and life-history. Here we perform a test of the hypothesized trade-off between the number and size of offspring using thousands of descriptions of the number of egg-producing compartments in the insect ovary (ovarioles), a common proxy for potential offspring number in insects. We find evidence of a negative relationship between egg size and ovariole number when accounting for adult body size. However in contrast to prior claims, we note that this relationship is not generalizable across all insect clades, and we highlight several factors that may have contributed to this size-number trade-off being stated as a general rule in previous studies. We reconstruct the evolution of the arrangement of cells that contribute nutrients and patterning information during oogenesis (nurse cells), and show that the diversification of ovariole number and egg size have both been largely independent of their presence or position within the ovariole. Instead we show that ovariole number evolution has been shaped by a series of transitions between variable and invariant states, with multiple independent lineages evolving to have almost no variation in ovariole number. We highlight the implications of these invariant lineages on our understanding of the specification of ovariole number during development, as well as the importance of considering developmental processes in theories of life-history evolution.

@@

The insect female reproductive system includes a pair of ovaries, each of which contains a number of ovarioles[@buning1994insect] (Fig 1a). Each ovariole consists of an anterior germarium containing the stem cell niche or resting oogonia, developing oocytes arranged in an ontogenic series from anterior to posterior, and a posterior connection to a common oviduct. The number of ovarioles varies across species[@hodin2009she], and can vary across individuals in a population[@honvek1993intraspecific], as well as between the left and right ovary within a single individual[@iwata1955comparative]. Therefore total ovariole number may be an even or odd integer for an individual female insect. In addition to variation in the number of ovarioles, the tissue morphology within ovarioles varies across insects, and has been classified into several modes of oogenesis based on the presence and position of special oocyte-associated  cells called nurse cells[@buning1994insect].

Here we compiled `r nrow(raw_ON)` records of ovariole number from across `r ON %>% distinct(order) %>% nrow()` orders, `r ON %>% distinct(family) %>% nrow()` families, and `r ON %>% filter(species != "") %>% distinct(genus,species) %>% nrow()` species of insects. We combined these data with published datasets of egg size[@church2019dataset], fecundity[@Gilbert2007;@gilbert2010parental], and body size[@rainford2016phylogenetic], to test hypotheses about the evolutionary trade-off between offspring size and number. In these analyses we used an existing phylogeny of insects[@church2019insect] to analyze evolutionary patterns in ovariole number, and found that hypotheses about life-history evolution do not hold generally true across insects. We then combined these data with published observations of the mode of oogenesis[@buning1994insect] and reconstructed the evolutionary history of the presence and position of nurse cells that contribute to the oocyte during oogenesis. We tested whether patterns in the distribution of ovariole number, egg size, or egg shape were driven by the evolution of nurse cells, and found no significant results. Instead we observe that the phylogenetic distribution of ovariole number suggests a model where the developmental mechanisms that govern ovariole number have shifted between variable and invariant states several times over the course of insect evolution. Based on this finding, we propose that the developmental mechanisms used to establish ovariole number in well-studied insects such as _Drosophila melanogaster_ are unlikely to regulate ovariole number in all insects.

@@

We searched the published literature for references to insect ovariole number using a predetermined set of 131 search terms, entered into Google Scholar (scholar.google.com) between June and October of 2019. Each search term was comprised of an insect taxonomic group and the words "ovariole number". The taxonomic groups used in the search process included all insect orders, many large insect families, and taxonomic groups that are well-represented in the insect egg dataset[@church2019dataset]. For each Google Scholar search, we evaluated the first ten pages in the search results. For 61 search terms that had a large number of informative hits, significant representation in the egg dataset, or that corresponded to very speciose groups, we evaluated an additional 20 publications. The list of search terms is available in the supplementary file 'ovariole_number_search_terms.tsv'.

Using this approach, we gathered `r nrow(raw_ON)` records for ovariole number for  `r ON %>% distinct(order) %>% nrow()` insect orders, `r ON %>% distinct(family) %>% nrow()` families, and `r ON %>% filter(species != "") %>% distinct(genus,species) %>% nrow()` species, using `r raw_ON %>% distinct(publication) %>% nrow()` publications that are listed in the supplementary file 'ovariole_number_bibliography.pdf'. We matched these records to additional taxonomic information using the software TaxReformer[@TaxReformer]. For all subsequent analyses, we excluded observations made in non-reproductive individuals from eusocial species (e.g. workers), as well as two observations that represented significant outliers and could not be validated using additional sources[@su2015testicular;@hernandez2017biology]. See supplementary methods section \ref{sec:gathering} for details.

For records of ovariole number that reported intraspecific variation in ovariole number, we calculated the percent difference as follows: if ovariole number was reported as a range, percent difference was calculated as the $100 * ((max - min) / median)$; if ovariole number was reported as an average with deviations, percent difference was calculated as $100 * ((2 * deviation) / mean)$. When independent observations of ovariole number for a given species were available across multiple published records, we calculated the percent difference as the $100 * ((max - min) / median)$.

@@

The analyses in this manuscript were performed using the insect phylogeny published in Church et al., 2019[@church2019insect], unless otherwise specified. For regressions involving body size data that were reported as insect family-level averages, we used the insect phylogeny published in Rainford et. al, 2014[@rainford2014phylogenetic]. Analyses of Drosophilidae ovariole number, egg size, and body size were performed using a phylogeny newly assembled for this study. See supplementary methods section \ref{sec:trees} for details.

@@

Analyses of evolutionary rate were performed using BAMM (version 2.5.0)[@rabosky2014automatic]. For this analysis, we calculated the average ovariole number (log~10~ transformed) for each genus present in the phylogeny (507 taxa). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select appropriate priors, and ran BAMM for the maximum number of generations ($2*10^{-9}$), sampling every 10^6^ generations. Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically. Running BAMM for the maximum possible number of generations and selecting the optimum burn-in (Fig. \ref{fig:BAMM-analysis-burnin}) resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for the log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See supplementary methods section \ref{sec:BAMM} for details.

@@

All phylogenetic regressions were performed using the maximum clade credibility (MCC) tree (the tree with highest credibility score from the posterior distribution of the Bayesian analysis). We considered a relationship significant when the p-value was below the threshold 0.01. To assess the robustness of results to uncertainty in phylogenetic relationships, we also repeated these analyses over the posterior distribution of phylogenetic trees and report the number of regressions that gave a significant result (see Table \ref{tab:PGLS-posterior-egg-ON}).

For two comparisons, we validated that our tests had sufficient statistical power using the selected threshold by comparing the distribution of p-values from regressions of observed data to regressions of data simulated under alternative hypotheses. We compared the results of analyses of our observed to those based on simulated data to evaluate the likelihood of false positives (comparing to data simulated under no correlation) and false negatives (comparing to data simulated with strong correlation).

@@

Ovariole number varies by at least four orders of magnitude across insect species (Fig. 1b). We identify seven insect families with species that have been reported to have more than 1,000 total ovarioles, including several eusocial insects (e.g. queens of the termite species _Hypotermes obscuriceps_, Blattodea: Termitidae[@bugnion1912anatomie], and several ant species, Hymenoptera: Formicidae)[@robertson1995sperm;@schneirla1957comparison] and non-eusocial insects (e.g. the blister beetle _Meloe proscarabaeus_, Coleoptera: Meloidae)[@buning1979trophic]. We also find two independent lineages that have evolved to have only one functional ovariole: dung beetles in the tribe Scarabaeinae (Coleoptera: Scarabaeidae)[@richter1974ovariole], and grass flies in the genus _Pachylophus_ (Diptera: Chloropidae)[@meier1999ovoviviparity;@pollock1996viviparous]. In these insects one of the two ovaries presumably established during embryogenesis is reported to atrophy during development[@pollock1996viviparous;@pluot1979evolution], resulting in an asymmetric adult reproductive system. We also evaluated intraspecific variation in ovariole number, and found that, for species for which it has been reported, the average percent difference number within species is between 10\% and 100\% of the median value (Fig. \ref{fig:intraspecific-variation}).

@@

Ovariole number has been hypothesized to be negatively correlated with egg size[@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. This hypothesis is based on the predictions that (1) female reproduction is resource-limited, therefore egg size should trade off with egg number, and (2) ovariole number can serve as a proxy for egg number[@montague1981reproductive;@smith1974optimal]. We did not observe a significant negative relationship when comparing egg size and ovariole number across insect species (Fig. 2a, Table \ref{tab:PGLS-posterior-egg-ON}, p-value `r round(pgls_results_egg_ON_name$tTable[[8]],3)`, n=`r pgls_results_egg_ON_name$dims$N`). We also compared egg size and ovariole number, combining data from species within the same genus to increase sample size, and again did not observe a significant relationship (Fig. \ref{fig:PGLS-egg-ON-genus}, p-value `r round(pgls_results_egg_ON_genus$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus$dims$N`). To verify this finding was not driven by the high ovariole numbers seen in the queens of some eusocial insects, we repeated this comparison excluding insects from families with eusocial representatives, with the same result (Fig. \ref{fig:PGLS-egg-ON-genus-noneusocial}, p-value `r round(pgls_results_egg_ON_genus_noneusocial$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus_noneusocial$dims$N`).

Given that this predicted relationship is often conditioned on body size, which is predicted to limit total potential reproductive investment[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with data on insect adult body mass[@Gilbert2007;@gilbert2010parental;@gilbert2011insect] and length[@rainford2016phylogenetic]. When accounting for adult body mass, we observed a significant negative relationship between egg size and ovariole number across genera (Fig. 2b, \ref{fig:PGLS-Gilbert-name}, p-value `r round(pgls_results_gilbert_genus$tTable[[8]],3)`, slope `r round(pgls_results_gilbert_genus$coefficients[[2]],3)`, n=`r pgls_results_gilbert_genus$dims$N`). To evaluate the robustness of this result, we repeated the analysis 1000 times, taking into account uncertainty in both the phylogeny and trait measurements. Out of 1000 regressions, `r sum(sapply(allometry_gilbert_genus,function(x){x$tTable[[8]]}) < 0.01)` indicated a significant negative relationship (Table \ref{tab:PGLS-posterior-egg-ON}). We performed the same comparison accounting for adult body length, and likewise observed a significant negative relationship (Fig. \ref{fig:PGLS-family-length}, p-value `r gsub(0,"<0.001",round(pgls_results_body_egg_ON_family$tTable[[8]],3))`, slope `r round(pgls_results_body_egg_ON_family$coefficients[[2]],3)`, n=`r pgls_results_body_egg_ON_family$dims$N`), supported by `r sum(sapply(allometry_body_egg_ON_family,function(x){x$tTable[[8]]}) < 0.01)` of 1000 repeated analyses (Table \ref{tab:PGLS-posterior-egg-ON}).

We further explored these results using two methods: First, to evaluate our findings against alternative evolutionary hypotheses, we compared these results to regressions based on simulated data. Our results showed that when considering body size, the slope of the regression of egg size and ovariole number is more negative than we would expect to observe by chance, as assessed by comparing to data simulated with no evolutionary correlation (Fig. \ref{fig:simulated}). However, for both adult body length and dry mass, the slope of the regressions on observed data are not within the range that would be expected under a strong negative correlation (slope of -1 in log-log space, Fig. \ref{fig:simulated}). This suggests the presence of a weak evolutionary relationship between ovariole number and egg size, when accounting for body size.

@@

In addition to the number of ovarioles, insect ovary morphology has been classified into several modes of oogenesis based on the presence and position of cells that provide nutritive and patterning molecules to the oocyte, which are called nurse cells[@buning1994insect] (Fig. 4a). Egg formation in the well-studied species _D. melanogaster_ is an example of a meroistic oogenesis mode, meaning that its ovarioles contain nurse cells of germ line origin that are connected to developing oocytes via cytoplasmic bridges[@KingRobertC1970OdiD]. In insects with a polytrophic meroistic arrangement, these nurse cells are clonally related and immediately adjacent to each oocyte. An alternative arrangement is seen in telotrophic meroistic ovaries, where oocytes in each ovariole are connected to a common pool of nurse cells located in the germarium[@buning1994insect]. Meroistic ovaries are thought to have evolved from an ancestral panoistic mode, meaning they lack nurse cells[@buning1994insect]. Using a previously published set of descriptions of these oogenesis modes across insects[@buning1994insect], we reconstructed the evolutionary transitions between these states. Consistent with previous analyses[@buning1994insect], we found that the ancestral insect likely had panoistic ovaries (lacking nurse cells), with several independent shifts to both telotrophic and polytrophic meroistic modes, and at least two reversals from meroistic back to panoistic (Figs. 4b, \ref{fig:reconstruct-ovary}).

@@

Using the dataset compiled here and a previously published phylogeny of insects (Fig. 5a)[@church2019insect], we modeled the rate of evolutionary change in ovariole number (Figs. \ref{fig:arbutus}, \ref{fig:BAMM-analysis-convergence}, \ref{fig:BAMM-analysis-burnin}, \ref{fig:BAMM-analysis-results}). We observed substantial rate heterogeneity in the evolution of ovariole number (Fig. \ref{fig:BAMM-analysis-results}), meaning that for some lineages ovariole number has evolved rapidly where in others, ovariole number has evolved very slowly or not at all. The most striking example of this are the multiple lineages which have independently evolved invariant or near-invariant ovariole number across taxa (e.g. nearly all Lepidoptera have exactly eight ovarioles, Fig. 5b, Lepidoptera are part of Amphiesmenoptera, in cyan), from an ancestral variable state. These invariant lineages were identified by finding regions of the phylogeny that experience extremely low rates of ovariole number diversification (Figs. \ref{fig:BAMM-analysis-results}, \ref{fig:BAMM-anlaysis-mean-rates}). Using this approach, we found that invariant ovariole numbers have evolved at least nine times independently across insects, with several subsequent reversals from invariant to variable states (Fig. 5a).

@@

The distribution of ovariole numbers across insects is enriched for even numbers of total ovarioles (Fig. 5c). While many insects show asymmetries in the number of ovarioles between the left and right ovaries, all of the invariant lineages are symmetric (at 4, 6, 8, 10, 12, and 14 total ovarioles). Additionally, for the insects identified as part of invariant lineages, none have any reported intraspecific variation in ovariole number. Therefore, invariant lineages have near-zero variation when comparing between species, between individuals within a species, and between the left and right ovary within an individual.

Using these results, we propose a multi-rate model, where the rate of ovariole number evolution differs based on the evolution of a discrete trait representing invariant or variable status. We propose that the evolution of this discrete trait is governed by a model where the likelihood of transitions from a variable to an invariant state is negatively correlated with the current number of ovarioles. Here we demonstrate that a multi-rate Brownian motion model far outperforms a single rate model in fitting the data ($\Delta$ AICc `r round(BM1$AICc - BMS$AICc,2)`). In addition, using a parametric bootstrap to evaluate model fit, we find evidence that processes beyond Brownian Motion processes are likely at play (Fig. \ref{fig:arbutus})[@pennell2015model]. We suggest that as researchers continue to develop non-Gaussian models for continuous trait evolution[@blomberg2020beyond], those models will be useful for describing the evolution of ovariole number.

@@

A frequently invoked life-history prediction is that, given a finite set of metabolic resources, organisms can either produce few offspring, each with high fitness, or many low-fitness offspring[@smith1974optimal;@lack1947significance;@stearns1992evolution]. In insects, egg size and ovariole number are often used as proxies for offspring fitness[@koch2014mother] and number[@david1970nombre;@bouletreau1982reproductive], respectively, and therefore it has been predicted that insects with more ovarioles lay smaller eggs than insects with fewer ovarioles[@hodin2009she;@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. Our results, using a dataset that spans `r nrow(raw_ON)` observations across `r ON %>% filter(species != "") %>% distinct(genus,species) %>% nrow()` species, and that takes into account phylogenetic relationships, indicate that a generalized trade-off between insect egg size and ovariole number does not exist (Fig. 2).

@@

We suggest that considering the evolution of developmental processes that govern ovariole number specification may be more useful in explaining patterns of diversity than predictions based on metabolic trade-offs. As evidence of this, we point to the fact that invariant lineages appear to have near-zero variation not only across species, but also within species, and between the left and right ovary within individuals. This suggests that the mechanism that determines ovariole number has become canalized in these groups. In contrast, our previous understanding of how ovariole number is regulated comes from research on _Drosophila melanogaster_, where the number of ovarioles can vary between the left and right ovaries within an individual, as well as across individuals within a population[@telonis2005genetic;@bergland2008quantitative]. In this species, adult ovariole number is determined by cell proliferation and rearrangement during larval development[@godt1995mechanisms;@king1968development]. Variation in adult number is derived primarily from variation in the number of “terminal filament precursor cells”[@green2012convergent;@sarikaya2019reproductive], as well as from variation in the number of those precursor cells that group together to form the structure that initiates ovariole formation, known as a “terminal filament”[@sarikaya2012roles]. Across species of _Drosophila_, variation in average adult ovariole number results primarily from variation in the average number of terminal filament precursor cells[@sarikaya2019reproductive].
